D3064R0: How to Avoid OOTA Without Really Trying

### 2024-02-24 ootavacuous.tex
Apply Alan Stern February 22 feedback
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

The C++ standard imposes many constraints, and these have been considered
in prior work.
However, in correct real-world C++ implementations, the standard's
abstract machine must be mapped onto real matter and energy.
This mapping has not been considered by prior work, and brings to
bear additional constraints from from hardware architecture and
design,\footnote{
	Which is in turn constrained by the laws of physics.}
which has been only partially accounted for in the standard and in
prior work.
One reason for the incomplete accounting for hardware constraints is
that accurate and executable formal descriptions of hardware memory
models did not appear until after the standard was released.

@@

Computer systems are expected to continue increasing their use of
speculative execution.
Nevertheless, the main points of this paper will remain unaffected.
To see this, keep in mind that on real systems, observable effects
must eventually be committed, but that nothing can be committed
until all of the computations that determine those effects have
committed.
Furthermore, instructions take time to execute, even when executed
speculatively.
For example, in Listing~\ref{lst:Speculated Store and Non-Speculated Load},
suppose that the storage for \co{x} is located immediately before that
for \co{y[]}.
Suppose further that hardware speculation incorrectly guesses the value
of \co{i} to be -1, so that the speculated store of 42 on line~9 uses
the address of \co{x}.\footnote{
	Yes, \co{y[-1]} is UB, but the CPU neither knows
	nor cares, nor should it.}
Then line~11 can cause the value 42 to be loaded into \co{r1}, when
by the rules of the C++ abstract machine the value must instead be
either zero or one.

@@

Because volatile atomics constitute observable
behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2023N4950},
they must be executed in strict accordance with the rules of the abstract
machine.
This means that any volatile atomic operation involving a store may
execute only after the committing of all loads whose return values are
used to compute that store's address or value.
OOTA afficienados will recognize this as a special case of ordering
relaxed loads before relaxed stores, albeit one not requiring
expensive memory-fence instructions on weakly ordered architectures.

@@

\paragraph{Invented Atomic Loads}
Although one can argue that it is functionally correct to invent atomic
loads as long as the loaded value is not used, doing so can result in
excessive cache-miss overheads.

@@

However, even duplicating an existing atomic load can lead to incorrect
results.
To see this, please refer to
Listing~\ref{lst:Do Not Duplicate Atomic Loads}.
Suppose that the load on line~13 were duplicated, for example, by
uncommenting line~15.
Then the values of \co{a} and \co{b} passed to line~17's call to
\co{do_something_with()} might be from different instances of the
\co{foo} structure, something that \co{do_something_with()} might
not be prepared to deal with.

Therefore, C++ implementations should avoid inventing atomic loads,
and in fact should avoid even duplicating them.

>>>>

The C++ standard imposes many constraints, and these have been considered
in prior work.
However, in correct real-world C++ implementations, the standard's
abstract machine must be mapped onto real matter and energy.
This mapping has not been considered by prior work, and brings to
bear additional constraints from hardware architecture and
design,\footnote{
	Which is in turn constrained by the laws of physics.}
which has been only partially accounted for in the standard and in
prior work.
One reason for the incomplete accounting for hardware constraints is
that accurate and executable formal descriptions of hardware memory
models did not appear until after the standard was released.

@@

Computer systems are expected to continue increasing their use of
speculative execution.
Nevertheless, the main points of this paper will remain unaffected.
To see this, keep in mind that on real systems, observable effects
must eventually be committed, but that no effect can be committed
while any of the computations that determine the effect remain
speculative.
Furthermore, instructions take time to execute, even when executed
speculatively.
For example, in Listing~\ref{lst:Speculated Store and Non-Speculated Load},
suppose that the storage for \co{x} is located immediately before that
for \co{y[]}.
Suppose further that hardware speculation incorrectly guesses the value
of \co{i} to be -1, so that the speculated store of 42 on line~9 uses
the address of \co{x}.\footnote{
	Yes, \co{y[-1]} is UB, but the CPU neither knows
	nor cares, nor should it.}
Then line~11 can cause the value 42 to be loaded into \co{r1}, when
by the rules of the C++ abstract machine the value must instead be
either zero or one.

@@

Because volatile atomics constitute observable
behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2023N4950},
they must be executed in strict accordance with the rules of the abstract
machine.
This means that any volatile atomic operation involving a store may
execute only after after all loads whose values are used to compute the
store's address or value have committed.
OOTA afficienados will recognize this as a special case of ordering
relaxed loads before relaxed stores, albeit one not requiring
expensive memory-fence instructions on weakly ordered architectures.

@@

\paragraph{Invented Atomic Loads}
Although one can argue that it is functionally correct to invent atomic
loads as long as the loaded value is not used, doing so can result in
excessive cache-miss overheads.
Even duplicating an atomic load can lead to incorrect results if done
carelessly.
To see this, please refer to
Listing~\ref{lst:Do Not Duplicate Atomic Loads}.
Suppose that the load on line~13 were duplicated, for example, by
uncommenting line~15.
Then the values of \co{a} and \co{b} passed to line~17's call to
\co{do_something_with()} might be from different instances of the
\co{foo} structure, something that \co{do_something_with()} might
not be prepared to deal with.

@@

For these reasons, C++ implementations should avoid inventing atomic
loads, and in fact should avoid even duplicating them.
